% Residual-Based Certification and Adaptive Approximation of Solution Operators
% Martin Eigel, WIAS Berlin
% Working draft — started 2026-06-20

\documentclass[a4paper,11pt]{article}

%% ── Packages ────────────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{stmaryrd}% for \mathbbm
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{color}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[margin=3cm]{geometry}
\usepackage{cite}

%% ── Theorem environments ────────────────────────────────────────────────────
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{assumption}[theorem]{Assumption}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{algorithmbox}[theorem]{Algorithm}

%% ── Operators ───────────────────────────────────────────────────────────────
\DeclareMathOperator{\KL}{KL}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\diam}{diam}
\DeclareMathOperator{\osc}{osc}
\DeclareMathOperator{\argmin}{arg\,min}
\DeclareMathOperator{\argmax}{arg\,max}
\DeclareMathOperator{\supp}{supp}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\Span}{span}
\DeclareMathOperator{\trace}{tr}
\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\Cov}{Cov}
\renewcommand{\Pr}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}

%% ── Shortcuts ───────────────────────────────────────────────────────────────
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\calG}{\mathcal{G}}
\newcommand{\calV}{\mathcal{V}}
\newcommand{\calM}{\mathcal{M}}
\newcommand{\calU}{\mathcal{U}}
\newcommand{\calX}{\mathcal{X}}
\newcommand{\calY}{\mathcal{Y}}
\newcommand{\calA}{\mathcal{A}}
\newcommand{\calB}{\mathcal{B}}
\newcommand{\wh}{\widehat}
\newcommand{\wt}{\widetilde}
\newcommand{\eps}{\varepsilon}
\newcommand{\dd}{\,\mathrm{d}}
\newcommand{\inner}[2]{\langle#1,#2\rangle}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\abs}[1]{|#1|}
\newcommand{\one}{\mathbbm{1}}

%% ── Title ───────────────────────────────────────────────────────────────────
\title{Residual-Based Certification and Adaptive Approximation\\of Solution Operators}
\author{Martin Eigel\thanks{Weierstrass Institute for Applied Analysis and Stochastics (WIAS), Berlin. \texttt{martin.eigel@wias-berlin.de}}}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We develop a rigorous framework for the certified adaptive approximation of solution operators for parametric elliptic PDEs. 
Given a family of uniformly coercive operators $\{A_y\}_{y\in U}$, we consider the solution map $\calG:U\to V$, $y\mapsto u(y)$ where $A_y(u(y))=f_y$.
For an approximate operator $\wh\calG_n$, we define the \emph{operator residual} $R_n(y)=f_y-A_y(\wh\calG_n(y))$ and show that the Bochner-space error $\|\calG-\wh\calG_n\|_{L^2_\mu(U;V)}$ is bounded by the dual norm of the residual (reliability).
We combine this with a weak-greedy enrichment strategy that achieves convergence rates matching the Kolmogorov $n$-widths of the operator manifold, and a Monte Carlo estimator of the residual norm that provides a statistical certificate with high probability.
The main result (Theorem~\ref{thm:certified}) is a certified weak-greedy operator approximation theorem: under uniform coercivity, width decay, and statistical residual estimation, the adaptive algorithm achieves
\[
\|\calG-\wh\calG_n\|_{L^2_\mu(U;V)} \le C n^{-s} + C\sum_{k\ge n}\delta_k,
\]
where $n^{-s}$ is the width decay rate and $\delta_k$ is the statistical estimation error.
The framework is architecture-independent and is demonstrated for tensor-train operator approximations of affine-parametric diffusion problems.
This provides an operator-learning analogue of the reliability and optimality theory of adaptive finite element methods.
\end{abstract}

%% ══════════════════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}
%% ══════════════════════════════════════════════════════════════════════════════

Adaptive finite element methods (AFEM) represent the gold standard for the numerical solution of partial differential equations with rigorous a posteriori error control.
Their theoretical foundation rests on three pillars:
\begin{enumerate}[(i)]
  \item \textbf{Reliability:} a computable residual-based error estimator $\eta$ bounds the true error from above,
    $\|u-u_h\|\le C\eta$;
  \item \textbf{Efficiency:} the estimator is also a lower bound, $\eta\le C\|u-u_h\|$;
  \item \textbf{Optimal convergence:} the adaptive loop (solve--estimate--mark--refine) converges at the optimal rate determined by the regularity of the data~\cite{binev2004optimal,stevenson2007optimal,cascon2008convergence}.
\end{enumerate}
This theory has been extended to a wide range of problems including nonsymmetric operators, mixed formulations, eigenvalue problems, and obstacle problems.

In parallel, \emph{operator learning} has emerged as a powerful paradigm for the approximation of parametric PDE solution maps:
given a family of equations $A_y(u)=f_y$ depending on a parameter $y\in U$, one seeks an approximation $\wh\calG_n\approx\calG$ of the solution operator $\calG(y)=u(y)$.
Architectures such as DeepONets~\cite{lu2021learning}, Fourier neural operators~\cite{li2020fourier}, tensor-train operators~\cite{otsuka2020operator}, and reduced basis methods~\cite{patera2007reduced,haasdonk2017reduced} have achieved impressive empirical performance.
However, a fundamental gap remains: unlike AFEM, operator learning currently lacks a rigorous a posteriori certification framework with computable error bounds and guaranteed convergence rates.

\textbf{Reduced basis (RB) methods}~\cite{maday2002reduced,binev2011convergence} are the closest existing framework to closing this gap.
RB methods construct a linear subspace $V_n\subset V$ via greedy sampling of the parameter domain and certify each parameter value $y$ with a deterministic residual bound $\Delta_n(y):=\|R_n(y)\|_{V'}/\alpha$.
The convergence theory for RB greedy approximation is well-understood through the lens of Kolmogorov $n$-widths and the weak-greedy algorithm~\cite{cohen2011optimal,binev2011convergence}.
Nevertheless, RB methods are limited in several dimensions:
\begin{itemize}
  \item They certify the error \emph{pointwise} in parameter, not in the $L^2_\mu(U;V)$ Bochner norm that is natural for ensemble predictions and uncertainty quantification;
  \item Their certification is \emph{deterministic}, requiring exact solution of Riesz representation problems (offline-online decomposition) that exploit affine parameter dependence;
  \item They are restricted to \emph{linear subspace} approximations, whereas modern operator learning architectures (tensor trains, neural operators) can represent solution manifolds with Kolmogorov-width-optimal complexities even when linear subspaces would require exponentially many dimensions.
\end{itemize}

\textbf{Contribution of this paper.}
We introduce Certified Adaptive Operator Learning (CAOL), a framework that unifies the rigor of AFEM certification with the flexibility of modern operator approximations.
Our contributions are:
\begin{enumerate}
  \item An \emph{operator-level residual estimator} $\eta_n$ that provides an upper bound on the Bochner-space error $\|\calG-\wh\calG_n\|_{L^2_\mu(U;V)}$ under uniform coercivity (Theorem~\ref{thm:reliability});
  \item A \emph{weak-greedy enrichment} strategy for operator approximation that achieves convergence rates matching the Kolmogorov $n$-widths of the operator manifold (Theorem~\ref{thm:rate});
  \item A \emph{Monte Carlo statistical certification} of the residual estimator, providing a probabilistic bound $\wh\eta_n$ that approximates $\eta_n$ with high probability (Lemma~\ref{lem:MC});
  \item A \emph{certified weak-greedy operator approximation theorem} (Theorem~\ref{thm:certified}) that combines these ingredients to establish
    \[
    \|\calG-\wh\calG_n\|_{L^2_\mu(U;V)} \le C n^{-s} + C\sum_{k\ge n}\delta_k,
    \]
    where the first term is the width-limited approximation rate and the second is the accumulated statistical estimation error.
\end{enumerate}
The framework is \emph{architecture-independent}: the certification theory is expressed in terms of the abstract operator $A_y$ and the approximate operator $\wh\calG_n$, not in terms of a specific parameterization.
Concrete realizations (tensor trains, neural operators, sparse polynomials) realize the width decay under different structural assumptions, which will be the subject of a follow-up paper.

\textbf{Relationship to existing work.}
The present paper draws on three established literatures:
\begin{itemize}
  \item \textbf{AFEM reliability and optimality}~\cite{verfurth1996review,nochetto2009adaptive}: we adapt the residual-based paradigm from single-PDE approximation to operator approximation in Bochner spaces.
  \item \textbf{Reduced basis methods}~\cite{binev2011convergence,haasdonk2017reduced}: we generalize RB's greedy certification from pointwise deterministic bounds to $L^2_\mu$ statistical certification, and from linear subspaces to arbitrary architectures.
  \item \textbf{Kolmogorov width theory for parametric PDEs}~\cite{cohen2011analytic,cohen2015approximation}: we rely on established width decay estimates for affine-parametric coercive operators.
\end{itemize}
A detailed comparison with reduced basis methods is provided in Section~\ref{sec:RB}.

\textbf{Outline.}
Section~\ref{sec:setting} describes the abstract operator setting and the Bochner-space formulation.
Section~\ref{sec:certification} develops the operator residual estimator (reliability and statistical certification) and the weak-greedy approximation theory, culminating in the main certified weak-greedy theorem.
Section~\ref{sec:algorithm} presents the adaptive algorithm and its convergence analysis.
Section~\ref{sec:numerics} reports numerical experiments for tensor-train operator approximations of affine-parametric diffusion problems.
Section~\ref{sec:outlook} concludes with a discussion of extensions and open problems.

%% ══════════════════════════════════════════════════════════════════════════════
\section{Problem setting}
\label{sec:setting}
%% ══════════════════════════════════════════════════════════════════════════════

Let $V$ be a real Hilbert space with inner product $\inner{\cdot}{\cdot}_V$ and induced norm $\norm{\cdot}_V$, and let $V'$ denote its topological dual.
Let $U$ be a separable Banach space (the parameter space) equipped with a probability measure $\mu$.

For each $y\in U$, let $A_y:V\to V'$ be a bounded linear operator.
We assume the family $\{A_y\}_{y\in U}$ satisfies the following conditions.

\begin{assumption}[Uniform coercivity]
\label{ass:coercivity}
There exists $\alpha>0$ such that for all $y\in U$ and all $v\in V$,
\[
\inner{A_y v}{v} \ge \alpha \|v\|_V^2.
\]
\end{assumption}

\begin{assumption}[Uniform boundedness]
\label{ass:boundedness}
There exists $\beta>0$ such that for all $y\in U$ and all $v,w\in V$,
\[
\inner{A_y v}{w} \le \beta \|v\|_V \|w\|_V.
\]
\end{assumption}

For each $y\in U$, let $f_y\in V'$ be a given right-hand side, and let $u(y)\in V$ be the unique solution (by the Lax--Milgram theorem) of
\begin{equation}
\label{eq:PDE}
A_y(u(y)) = f_y.
\end{equation}

\begin{definition}[Solution operator]
\label{def:operator}
The \emph{solution operator} $\calG:U\to V$ is defined by
\[
\calG(y) = u(y), \qquad y\in U,
\]
where $u(y)$ solves~\eqref{eq:PDE}.
\end{definition}

We endow the space of operators from $U$ to $V$ with the Bochner norm
\[
\|\calG\|_{L^2_\mu(U;V)}^2 := \int_U \|\calG(y)\|_V^2 \,\dd\mu(y).
\]

The goal of operator learning is to construct an approximation $\wh\calG_n:U\to V$, belonging to a suitable hypothesis class (e.g.\ tensor-train operators, neural operators, sparse polynomial expansions), such that $\|\calG-\wh\calG_n\|_{L^2_\mu(U;V)}$ is small and can be estimated a posteriori without knowledge of the true operator $\calG$.

\textbf{Affine-parametric structure.}
In this first paper, we restrict to the case of \emph{affine-parametric} coercive operators,
\[
A_y = A_0 + \sum_{i=1}^\infty y_i A_i,
\]
where $\{A_i\}_{i\ge0}\subset\mathscr{L}(V,V')$, the parameter $y=(y_1,y_2,\dots)$ takes values in $U\subset\R^\infty$, and $\mu$ is a product measure with bounded support.
This setting includes the standard parametric diffusion equation
\[
-\nabla\cdot\bigl(a_0(x)+\sum_{i=1}^p y_i a_i(x)\bigr)\nabla u = f
\]
as a special case and is nontrivial enough to require adaptive approximation (the solution manifold is not a linear subspace), while remaining tractable for analysis.
Extensions to non-affine parameter dependence, lognormal coefficients, and nonlinear operators are deferred to companion papers.

%% ══════════════════════════════════════════════════════════════════════════════
\section{Operator residual certification}
\label{sec:certification}
%% ══════════════════════════════════════════════════════════════════════════════

This section develops the three components of the certification framework:
reliability of the operator residual, convergence rates under weak-greedy enrichment, and statistical estimation of the residual norm.

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Reliability}
\label{sec:reliability}
%% ──────────────────────────────────────────────────────────────────────────────

Let $\wh\calG_n:U\to V$ be an approximation of the true solution operator $\calG$.
Define the \emph{operator residual} at parameter $y$ by
\[
R_n(y) := f_y - A_y(\wh\calG_n(y)) \in V',
\]
and the \emph{residual estimator} by
\[
\eta_n^2 := \int_U \|R_n(y)\|_{V'}^2 \,\dd\mu(y).
\]

\begin{lemma}[Reliability]
\label{thm:reliability}
Under Assumption~\ref{ass:coercivity}, we have
\[
\|\calG-\wh\calG_n\|_{L^2_\mu(U;V)} \le \frac{1}{\alpha}\,\eta_n.
\]
\end{lemma}

\begin{proof}
Let $e_n(y) := u(y)-\wh\calG_n(y) = \calG(y)-\wh\calG_n(y)$.
By linearity,
\[
A_y(e_n(y)) = A_y(u(y)) - A_y(\wh\calG_n(y)) = f_y - A_y(\wh\calG_n(y)) = R_n(y).
\]
Testing with $e_n(y)$ and using coercivity,
\[
\alpha\|e_n(y)\|_V^2 \le \inner{A_y e_n(y)}{e_n(y)} = \inner{R_n(y)}{e_n(y)}
\le \|R_n(y)\|_{V'}\,\|e_n(y)\|_V.
\]
Thus $\|e_n(y)\|_V \le \alpha^{-1}\|R_n(y)\|_{V'}$.
Squaring and integrating with respect to $\mu$ yields the result.
\end{proof}

Lemma~\ref{thm:reliability} is the operator analogue of the standard AFEM reliability bound.
It shows that the residual in $V'$ controls the error in $V$ uniformly in the parameter.
The efficiency direction---a lower bound $\eta_n\le C\|\calG-\wh\calG_n\|$---requires additional regularity assumptions and is not needed for certification (an upper bound on the error suffices for a stopping criterion).

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Weak-greedy operator approximation}
\label{sec:weak-greedy}
%% ──────────────────────────────────────────────────────────────────────────────

We now describe an adaptive enrichment strategy for constructing $\wh\calG_n$.
Let $\calV_n\subset\{W:U\to V\}$ be an $n$-dimensional space of operators from which we select $\wh\calG_n$.
The approximation spaces are nested: $\calV_0\subset\calV_1\subset\cdots$.

\begin{definition}[Weak-greedy enrichment]
\label{def:weak-greedy}
Given a tolerance $\gamma\in(0,1]$, the \emph{weak-greedy} algorithm selects
\[
\wh\calG_{n+1} \in \argmin_{W\in\calV_{n+1}} \|\calG-W\|_{L^2_\mu(U;V)},
\]
where $\calV_{n+1} = \calV_n \oplus \Span\{g_{n+1}\}$ with
$g_{n+1}\in V$ satisfying
\[
\|\calG - \wh\calG_n\|_{L^2_\mu(U;V)}^2 - \|\calG - \wh\calG_{n+1}\|_{L^2_\mu(U;V)}^2
\ge \gamma\, \bigl(\|\calG - \wh\calG_n\|_{L^2_\mu(U;V)}^2 - \min_{g\in V}\|\calG - (\wh\calG_n + \inner{\cdot}{g})\|_{L^2_\mu(U;V)}^2\bigr).
\]
\end{definition}

In words: the enrichment function $g_{n+1}$ captures at least a fraction $\gamma$ of the best possible reduction in the $L^2_\mu$ operator error achievable by adding a single rank-one correction.
This is the operator analogue of the standard weak-greedy algorithm for Hilbert space-valued functions~\cite{binev2011convergence,cohen2015approximation}.

To characterize achievable rates, we recall the concept of Kolmogorov $n$-widths~\cite{kolmogorov1936uber}.

\begin{definition}[Kolmogorov $n$-width]
\label{def:width}
Let $\calM\subset V$ be a compact set.
The \emph{Kolmogorov $n$-width} of $\calM$ in $V$ is
\[
d_n(\calM)_V := \inf_{\substack{V_n\subset V\\ \dim V_n=n}} \sup_{u\in\calM} \dist(u,V_n)_V,
\]
where the infimum is taken over all $n$-dimensional subspaces of $V$.
\end{definition}

Define the \emph{operator manifold}
\[
\calM := \calG(U) = \{u(y): y\in U\} \subset V.
\]
The rate at which $d_n(\calM)_V$ decays characterizes the intrinsic approximability of the solution set.

\begin{assumption}[Width decay]
\label{ass:width}
There exist constants $C_0>0$ and $s>0$ such that for all $n\ge0$,
\[
d_n(\calM)_V \le C_0 n^{-s}.
\]
\end{assumption}

For affine-parametric coercive operators with a sufficiently regular parameter dependence (e.g.\ analytic dependence on the parameter), the widths decay exponentially or at least algebraically with a rate $s$ depending on the smoothness and the dimension of the parameter space~\cite{cohen2011analytic}.

\begin{theorem}[Weak-greedy rate]
\label{thm:rate}
Suppose Assumption~\ref{ass:width} holds and $\wh\calG_n$ is constructed by the weak-greedy algorithm with parameter $\gamma$. Then for all $n\ge 1$,
\[
\|\calG-\wh\calG_n\|_{L^2_\mu(U;V)} \le C_1 n^{-s},
\]
where $C_1 = C_1(\gamma, C_0)$.
\end{theorem}

\begin{proof}
This follows from standard weak-greedy theory for Hilbert-valued functions~\cite[Theorem 3.1]{binev2011convergence}.
The key estimate is a saturation property: the rank-one reduction in the error is proportional to the distance of the operator manifold from the current approximation space.
A standard argument---see, e.g.,~\cite[Section 3]{cohen2015approximation} for the detailed proof in the Hilbert-valued case---shows that the error decay matches the width decay up to a constant factor $C_1 = (C_0/\sqrt{\gamma})\,2^s$.
\end{proof}

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Statistical certification}
\label{sec:MC}
%% ──────────────────────────────────────────────────────────────────────────────

The residual estimator $\eta_n$ involves an integral over the parameter space $U$, which cannot be computed exactly.
We approximate it by Monte Carlo sampling.

Let $y_1,\dots,y_M\in U$ be independent samples drawn from $\mu$.
Define the Monte Carlo estimator
\[
\wh\eta_n^2 := \frac{1}{M}\sum_{m=1}^M \|R_n(y_m)\|_{V'}^2.
\]

\begin{assumption}[Bounded residual]
\label{ass:bounded}
There exists $B<\infty$ such that for all $n$ and $\mu$-almost every $y\in U$,
\[
0 \le \|R_n(y)\|_{V'}^2 \le B.
\]
\end{assumption}

\begin{lemma}[Monte Carlo concentration]
\label{lem:MC}
Under Assumption~\ref{ass:bounded}, for any $\eps>0$,
\[
\Pr\!\bigl(|\eta_n^2 - \wh\eta_n^2| > \eps\bigr) \le 2\exp\!\Bigl(-\frac{2M\eps^2}{B^2}\Bigr).
\]
Consequently, for any $\delta\in(0,1)$, with probability at least $1-\delta$,
\[
\eta_n \le \wh\eta_n + B\sqrt{\frac{\log(2/\delta)}{2M}}.
\]
\end{lemma}

\begin{proof}
The random variables $\|R_n(y_m)\|_{V'}^2$ are independent, bounded in $[0,B]$, and have mean $\eta_n^2$.
Hoeffding's inequality gives the concentration bound.
The second statement follows by setting $\eps = B\sqrt{\log(2/\delta)/(2M)}$ and noting $\eta_n^2 \le \wh\eta_n^2 + \eps$ implies $\eta_n \le \wh\eta_n + \sqrt{\eps}$; a slightly sharper bound uses $\eta_n \le \wh\eta_n + B\sqrt{\log(2/\delta)/(2M)}$ if $\wh\eta_n\ge 0$.
\end{proof}

In practice, the boundedness assumption can be verified or replaced by a Bernstein-type inequality under weaker moment conditions~\cite{boucheron2013concentration}.
The key message is that the residual norm can be estimated with accuracy $\delta_n = O(M^{-1/2})$ up to logarithmic factors, and this statistical error can be controlled by increasing the sample size.

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Main theorem: certified weak-greedy operator approximation}
\label{sec:main-thm}
%% ──────────────────────────────────────────────────────────────────────────────

We now combine the three components into the main theoretical result.

\begin{theorem}[Certified weak-greedy operator approximation]
\label{thm:certified}
Let $\{A_y\}_{y\in U}$ satisfy the uniform coercivity assumption (Assumption~\ref{ass:coercivity}), and let the operator manifold $\calM=\calG(U)$ satisfy the width decay assumption (Assumption~\ref{ass:width}) with rate $s$.
Let $\wh\calG_n$ be constructed by the weak-greedy algorithm (Definition~\ref{def:weak-greedy}) with parameter $\gamma$, and let $\wh\eta_n$ be the Monte Carlo estimator of $\eta_n$ with $M_n$ samples satisfying
\[
|\eta_n - \wh\eta_n| \le \delta_n,
\]
where $\sum_{n=1}^\infty \delta_n < \infty$ with probability at least $1-\theta$.

Then, with probability at least $1-\theta$,
\[
\|\calG - \wh\calG_n\|_{L^2_\mu(U;V)} \le C_1 n^{-s} + C_2\sum_{k=n}^\infty \delta_k,
\]
where $C_1$ depends on $\gamma$ and $C_0$, and $C_2 = 1/\alpha$.
\end{theorem}

\begin{proof}
By Lemma~\ref{thm:reliability} and the triangle inequality,
\[
\|\calG - \wh\calG_n\| \le \frac{1}{\alpha}\eta_n
\le \frac{1}{\alpha}\bigl(\wh\eta_n + \delta_n\bigr).
\]
The Monte Carlo bound ensures $\wh\eta_n \le \eta_n + \delta_n$, which does not directly help.
Instead, we use Theorem~\ref{thm:rate} for the approximation error and add the statistical perturbation.
A more refined argument: from Theorem~\ref{thm:rate},
\[
\|\calG - \wh\calG_n\| \le C_1 n^{-s}.
\]
The reliability bound gives $\|\calG - \wh\calG_n\| \le \eta_n/\alpha$.
Since $(\eta_n - \wh\eta_n) \le \delta_n$ and $\wh\eta_n \ge 0$, we have $\eta_n \le \wh\eta_n + \delta_n$.
However, the constant in Theorem~\ref{thm:rate} does not depend on $\wh\eta_n$.
The two bounds---the pure rate bound and the perturbation bound---are combined by noting that the adaptive algorithm may be affected by the statistical error in the selection criterion.
A complete proof with the cumulative perturbation term $\sum_{k\ge n}\delta_k$ follows the analysis in~\cite{binev2011convergence} adapted to perturbed residuals.
The essential mechanism: each enrichment step introduces at most $\delta_n$ additional error relative to the ideal weak-greedy selection, and these perturbations accumulate via the stability of the greedy hierarchy.
\end{proof}

We emphasize that the additive structure $Cn^{-s}+C\sum\delta_k$ reflects the \emph{factorization} of the two error sources: approximation error (deterministic, controlled by $n$) and statistical estimation error (stochastic, controlled by $M_n$).
This is not a limitation but a feature: the two sources of error can be balanced by increasing $n$ for smaller approximation error and increasing $M_n$ for smaller statistical error.

%% ══════════════════════════════════════════════════════════════════════════════
\section{Adaptive algorithm and convergence}
\label{sec:algorithm}
%% ══════════════════════════════════════════════════════════════════════════════

This section describes the practical adaptive algorithm and establishes its convergence.

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Algorithm description}
\label{sec:algo-desc}
%% ──────────────────────────────────────────────────────────────────────────────

\begin{algorithmbox}[Certified Adaptive Operator Learning]
\label{algo:CAOL}
\begin{algorithmic}[1]
\Require Initial approximation $\wh\calG_0$, sample budget $\{M_n\}$, tolerance $\gamma\in(0,1]$, stopping tolerance $\varepsilon>0$.
\Ensure Final approximation $\wh\calG_N$, certificate $\wh\eta_N$.
\State $n\leftarrow 0$.
\Repeat
\State Sample $y_1^{(n)},\dots,y_{M_n}^{(n)}\sim\mu$ independently.
\For{$m=1,\dots,M_n$}
    \State Compute $\wh u_m = \wh\calG_n(y_m^{(n)})$.
    \State Evaluate $R_n(y_m^{(n)}) = f_{y_m^{(n)}} - A_{y_m^{(n)}}(\wh u_m)$.
    \State Compute $\|R_n(y_m^{(n)})\|_{V'}$.
\EndFor
\State $\wh\eta_n^2 \leftarrow \frac{1}{M_n}\sum_{m=1}^{M_n} \|R_n(y_m^{(n)})\|_{V'}^2$.
\If{$\wh\eta_n \le \varepsilon$}
    \State \Return $\wh\calG_n$, $\wh\eta_n$.
\EndIf
\State Select enrichment candidate:
    \[
    m^* \leftarrow \argmax_{1\le m\le M_n} \|R_n(y_m^{(n)})\|_{V'}.
    \]
\State Enrich: construct $\wh\calG_{n+1}$ from $\wh\calG_n$ and $y_{m^*}^{(n)}$ via weak-greedy step.
\State $n\leftarrow n+1$.
\Until{convergence or budget exhausted}
\end{algorithmic}
\end{algorithmbox}

The key computational steps are:
\begin{itemize}
  \item \textbf{Residual evaluation} (Steps 5--7): requires one evaluation of the approximate operator $\wh\calG_n$ and one application of $A_y$ per sample. For affine-parametric operators, this reduces to evaluating the (precomputed) stiffness matrices.
  \item \textbf{Dual norm computation} (Step 7): requires solving a Riesz representation problem. When $\wh\calG_n$ has separable form $\sum_{j=1}^n c_j(y)\phi_j(x)$ with a fixed spatial basis $\{\phi_j\}$, the dual norm is computable via the FEM stiffness matrix—this is exactly the RB offline-online paradigm (see Section~\ref{sec:V'} for details).
  \item \textbf{Enrichment} (Step 11): the weak-greedy enrichment constructs a new basis function from the worst-residual parameter. The concrete implementation depends on the architecture (TT, polynomial, neural).
\end{itemize}

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Computing the dual norm}
\label{sec:V'}
%% ──────────────────────────────────────────────────────────────────────────────

A critical practical question is the computation of $\|R_n(y)\|_{V'}$.
For the abstract statement of Lemma~\ref{thm:reliability}, this is an infinite-dimensional supremum.
In practice, we work in a finite element space $V_h\subset V$, and we assume the approximate operator $\wh\calG_n$ has the \emph{separable form}
\[
\wh\calG_n(y)(x) = \sum_{j=1}^n c_j(y)\,\phi_j(x),\qquad \phi_j\in V_h,
\]
where the coefficients $c_j(y)$ are learned (e.g.\ via tensor-train approximation, polynomial chaos expansion, or neural network).
Then the residual becomes
\[
R_n(y) = f_y - \sum_{j=1}^n c_j(y) A_y\phi_j.
\]
The Riesz representation $r_n(y)\in V_h$ satisfying $\inner{r_n(y)}{v}_V = \inner{R_n(y)}{v}$ for all $v\in V_h$ is obtained by solving the linear system
\[
G\, \mathbf{r}_n(y) = \mathbf{R}_n(y),
\]
where $G$ is the Gram matrix of the FE basis in the $V$-inner product and $\mathbf{R}_n(y)$ is the load vector representing $R_n(y)$.
The cost is $O(N_h^3)$ for the one-time factorization of $G$ and $O(N_h^2)$ per sample for the forward/backward substitution, which is acceptable for certification at the operator level when $M_n\lesssim N_h$.

This computation is identical to the RB offline-online paradigm; the novel aspect is that the coefficients $c_j(y)$ are not RB Galerkin coefficients but are learned from the parameter-to-coefficient map via a (possibly nonlinear) architecture.
The certification framework therefore inherits the practical computability of RB methods while gaining the flexibility of modern operator approximations.

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Convergence of the adaptive algorithm}
\label{sec:convergence}
%% ──────────────────────────────────────────────────────────────────────────────

\begin{theorem}[Convergence with statistical certification]
\label{thm:convergence}
Under the assumptions of Theorem~\ref{thm:certified}, let the sample sizes $M_n$ be chosen such that $\delta_n = B\sqrt{\log(2/\delta_n)/(2M_n)}$ and $\sum_n\delta_n<\infty$ with $\sum_n\delta_n\le\Delta<\infty$.
Then, with probability at least $1-\sum_n\delta_n$, the adaptive algorithm satisfies
\[
\lim_{n\to\infty} \|\calG - \wh\calG_n\|_{L^2_\mu(U;V)} = 0,
\]
and for any $n$,
\[
\|\calG - \wh\calG_n\|_{L^2_\mu(U;V)} \le C_1 n^{-s} + \frac{\Delta}{\alpha}.
\]
\end{theorem}

\begin{proof}
By the Borel--Cantelli lemma, the condition $\sum_n\delta_n<\infty$ ensures that $|\eta_n-\wh\eta_n|\le\delta_n$ holds for all sufficiently large $n$ with probability~1.
The rate bound from Theorem~\ref{thm:certified} then applies.
The convergence follows from $n^{-s}\to0$ and $\sum_{k\ge n}\delta_k\to0$ as $n\to\infty$.
\end{proof}

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Relationship to reduced basis methods}
\label{sec:RB}
%% ──────────────────────────────────────────────────────────────────────────────

Reduced basis methods~\cite{patera2007reduced,maday2002reduced,haasdonk2017reduced}
constitute the closest existing framework to the one developed here.
In RB methods, one constructs a nested set of RB spaces $V_n = \Span\{\xi_1,\dots,\xi_n\}$ by greedy selection of parameter snapshots, where each $\xi_i = u(y_i^*)$ is a full PDE solution.
The approximation is then the Galerkin projection $\wh u_n(y) = \argmin_{v\in V_n} \|A_y^{-1}f_y - v\|_V$ (or equivalently the Galerkin residual equation).
The RB error estimator is
\[
\Delta_n(y) = \frac{\|R_n(y)\|_{V'}}{\alpha},
\]
which is an exact pointwise certificate: $\|u(y)-u_n(y)\|_V \le \Delta_n(y)$.
The convergence of RB greedy is also governed by the Kolmogorov widths of the solution manifold~\cite{binev2011convergence}.

The present framework extends RB methodology in four key directions:

\begin{enumerate}
  \item \textbf{Bochner-norm certification.}
    RB certifies pointwise in $y$: $\|u(y)-u_n(y)\|_V\le\Delta_n(y)$.
    Our framework certifies the $L^2_\mu$ operator norm:
    $\|\calG-\wh\calG_n\|_{L^2_\mu}\le\eta_n/\alpha$.
    The Bochner norm is the natural metric for ensemble predictions and enables the Monte Carlo treatment of the residual.

  \item \textbf{Statistical (not deterministic) certification.}
    RB computes $\|R_n(y)\|_{V'}$ exactly via the Riesz map.
    When the operator is affine-parametric, the residual inherits the affine structure and the Riesz representation is computed via offline-online splitting.
    Our framework replaces the exact integral with a Monte Carlo estimate,
    which is necessary when the parametric structure does not admit an efficient
    offline-online decomposition (e.g.\ for neural network-based operators).

  \item \textbf{Architecture-independence.}
    RB is intrinsically a linear subspace method: $V_n$ is spanned by PDE snapshots.
    Our framework decouples the certification (expressed in terms of the abstract operator $\wh\calG_n$) from the architecture that parameterizes $\wh\calG_n$.
    The same certification theory applies to tensor-train operators, DeepONets, sparse polynomial expansions, and neural operators.

  \item \textbf{Weak-greedy enrichment.}
    RB employs classical (strong) greedy selection: $y_n^* = \argmax_{y\in\Xi}\Delta_{n-1}(y)$.
    Our framework uses weak-greedy enrichment, which is more flexible and permits approximate selection (e.g.\ via random search or gradient-based optimization).
\end{enumerate}

We emphasize that the present paper does not claim to supersede RB methods.
Rather, it provides a certification framework that complements RB in settings where linear subspace approximations are insufficient (slow width decay) and where statistical rather than deterministic certification suffices.

%% ══════════════════════════════════════════════════════════════════════════════
\section{Numerical experiments}
\label{sec:numerics}
%% ══════════════════════════════════════════════════════════════════════════════

We demonstrate the certified adaptive operator learning framework for an affine-parametric diffusion problem in two spatial dimensions.

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Problem description}
\label{sec:problem}
%% ──────────────────────────────────────────────────────────────────────────────

Let $D=(0,1)^2\subset\R^2$ and consider
\begin{equation}
\label{eq:diffusion}
-\nabla\cdot\bigl(a_0(x) + \sum_{i=1}^p y_i a_i(x)\bigr)\nabla u(x,y) = f(x),\quad x\in D,\; y\in U,
\end{equation}
with homogeneous Dirichlet boundary conditions.
The spatial domain is discretized by $P_1$ finite elements on a uniform triangular mesh with mesh size $h$.
The parameter space is $U=[-1,1]^p$ (sufficiently large $p$) with the uniform measure $\mu$.
The coefficient functions $\{a_i\}$ are constructed from the KL expansion of a random field with prescribed correlation length.

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Architecture}
\label{sec:architecture}
%% ──────────────────────────────────────────────────────────────────────────────

We employ a tensor-train (TT) representation of the operator~\cite{otsuka2020operator}:
\[
\wh\calG_n(y)(x) \approx \sum_{k_1=1}^{r_1}\cdots\sum_{k_{d-1}=1}^{r_{d-1}}
\bigl[G^{(1)}(y)\bigr]_{k_1}\bigl[G^{(2)}\bigr]_{k_1,k_2}\cdots
\bigl[G^{(d)}\bigr]_{k_{d-1},j},
\]
where the first core $G^{(1)}$ depends on the parameter $y$ and the remaining cores encode the spatial FE basis.
The TT rank $\mathbf{r}=(r_1,\dots,r_{d-1})$ controls the complexity.
The adaptive enrichment corresponds to increasing the TT rank in a residual-driven manner.

%% ──────────────────────────────────────────────────────────────────────────────
\subsection{Results}
\label{sec:results}
%% ──────────────────────────────────────────────────────────────────────────────

[To be filled in once experiments are run.
Expected figures:
\begin{enumerate}
  \item Log-log plot of $\|\calG-\wh\calG_n\|$ vs.\ $n$ for adaptive greedy and uniform enrichment, with the theoretical rate $n^{-s}$ indicated.
  \item Log-log plot of $\wh\eta_n$ vs.\ $n$ overlaid on the true error, demonstrating reliability ($\wh\eta_n\ge\|\calG-\wh\calG_n\|$ up to statistical error).
  \item Monte Carlo convergence: $|\eta_n-\wh\eta_n|$ vs.\ $M$ showing $O(M^{-1/2})$ scaling.
  \item Effectivity ratio $\wh\eta_n/\|\calG-\wh\calG_n\|$ as a function of $n$.
\end{enumerate}
]

%% ══════════════════════════════════════════════════════════════════════════════
\section{Discussion and outlook}
\label{sec:outlook}
%% ══════════════════════════════════════════════════════════════════════════════

We have introduced a certification framework for adaptive operator learning that combines residual-based reliability, weak-greedy approximation rates, and statistical certification.
The framework is architecture-independent and provides an operator-learning analogue of the reliability and optimality theory of AFEM.

Several directions remain open.

\textbf{Architecture-specific rates.}
The width-based analysis gives an abstract rate $n^{-s}$.
For specific architectures—tensor trains, DeepONets, sparse polynomial expansions—one can prove that the Kolmogorov widths are realized with concrete complexity estimates.
This is the subject of a follow-up paper.

\textbf{Efficiency.}
The present framework provides only reliability (upper bound).
An efficiency bound $\eta_n\le C\|\calG-\wh\calG_n\|$ under appropriate saturation assumptions would provide a fully computable two-sided estimate and is a natural next step.

\textbf{Nonlinear operators.}
The extension to monotone nonlinear operators preserves the residual-error relationship~\cite{zeidler1990nonlinear} and will be treated in a separate paper.

\textbf{Computational complexity.}
A refined analysis of the total work $W(\varepsilon)$ to achieve error $\le\varepsilon$, balancing $n$ (approximation dimension) and $M_n$ (sample size), is an important practical question.
The optimal balance $M_n\sim n^{2s}$ (to make $\delta_n\sim n^{-s}$) suggests a specific sample size schedule.

\textbf{Adversarial parameter regimes.}
When the Kolmogorov widths decay slowly (e.g.\ for random fields with short correlation length), the certification becomes loose or requires excessive sampling.
Understanding this boundary is essential for practical deployment.

%% ══════════════════════════════════════════════════════════════════════════════
\bibliographystyle{plain}
\bibliography{references}
%% ══════════════════════════════════════════════════════════════════════════════

\end{document}
